---
title: "ACOM - ACCORD Fire Workshop 2017"
author: "Nicholas Good - n.good@colostate.edu"
output:
  rmarkdown::html_document:
    toc: true
    toc_float: true
    theme: yeti
---

```{r global_options, include=FALSE}
  knitr::opts_chunk$set(echo=TRUE, warning=FALSE, message=FALSE, results = 'hide', eval = FALSE)
```

---

# Getting started

## Exercises

You can download the .html exercises file from here: [exercises](https://tinyurl.com/R-for-fire-data).

## R projects

You can create an *R Studio* project to organize your work. An **R Studio** project has its own working directory, workspace, history, and source files.

* To create an **R Studio** project navigate to: `File -> New Project -> Create Project` (giving the new folder a name via the dialog).

* Select either `New directory -> Empty Project` or `Existing Directory -> Create Project` (selecting an existing folder via the dialog)

## R packages

The base functionality of R is expanded upon by installing additional packages. In this section we'll install all the packages we're going to use today. You can install packages using the `install.packages()` function or via the **R Studio** user interface (`tools -> install packages`).

* The [rmarkdown](http://rmarkdown.rstudio.com) package allows you to combine: code, its output and text into a single document in **R Studio** (like the one you are reading now). **R Markdown** supports R and other languages (e.g. Python and SQL). You can output in numerous formats such as *html*, *word* and *pdf*. We'll output to **html**.

```{r,}
  install.packages("rmarkdown")
  install.packages("knitr")
```

* The [tidyverse](http://tidyverse.org) is a collection of [data science](http://r4ds.had.co.nz) focused packages underpinned by common (and evolving) approaches to data manipulation.

* The `install.packages("tidyverse")` command will install the core *tidyverse* packages

```{r,}
  install.packages("tidyverse")
```

* Some additional tidyverse packages we'll use are `magrittr` that adds operators to make code more readable and `lubridate` for date-time manipulation.

```{r,}
  install.packages("magrittr")
  install.packages("lubridate")
```

* The `tidyverse` includes the powerful `ggplot2` library. For spatial analysis we'll include install the `ggmaps` and `plotly` packages. We'll also install the `ggthemes` package.

```{r,}
  install.packages("ggmaps")
  install.packages("plotly")
  install.packages("ggthemes")
```

---

## Data

We're going to look at some data from the [Arctas](https://www-air.larc.nasa.gov/cgi-bin/arcstat-c) project. You'll need to download the [dataset](https://tinyurl.com/R-for-fire-data). Navigate to `data -> arctas` download the data and save it to a folder **inside your project folder** `project folder/arctas/`.

---

## Source file

To speed things up we'll use some code I've already written. You'll need to download the [scripts folder](https://tinyurl.com/R-for-fire-data). Download the folder and save it in your **project folder**.

---

# R Markdown

## Resources

* R Studio's [r markdown website](http://rmarkdown.rstudio.com/lesson-1.html)

* The [r markdown cheatsheet](https://www.rstudio.com/wp-content/uploads/2016/03/rmarkdown-cheatsheet-2.0.pdf)

* The [r markdown reference](https://www.rstudio.com/wp-content/uploads/2015/03/rmarkdown-reference.pdf)

---

## Creating a markdown document

* Navigate to `File -> New File -> R Markdown -> ok` making sure the `document` and `html` options are selected. You'll also want to give your project a name.

* Save your **R markdown** file in your project file (`File Save`).

* Press the **Knit** button, what happens?

* Take a moment to relate the text in your markdown document to the output.

* Remove all the code from the **R Markdown** heading down.

* Use the `#` symbol to create a new section header **Load Data**

* Press the **Knit** button again.

You'll use this file to document your code in the next steps.

---

## Code chunks

You can add a *chunk* of code to your *markdown* document using the *insert button*. 

* Add several carriage returns below your title and press `insert -> R`.

* Copy the code below into the body of the *chunk*.

* Use green arrow in the top right of the chunk box to run the code.

```{r, cache=TRUE, eval=TRUE}
  library(tidyverse)
  library(magrittr)

  source("scripts/r_functions.R")

  arctas_data <- lapply(list.files("../data/arctas",
                        pattern = "_R14.ict$",
                        full.names = TRUE),
                        read_ict_file) %>%
                 bind_rows() %>%
                 mutate_all(funs(na_if(., -999999999))) %>%
                 mutate_all(funs(na_if(., -888888888)))
```

* Take a look at your **Environment** tab, what do you see?

---

## Inline code

```{r, echo=FALSE, eval=TRUE}
 sourcecode <- "\\`r code here\\`"
 sourcecode2 <- "\\`r length(unique(arctas_data$flight))\\`"
```

You can write code that is evaluated inline with your text using `r sourcecode`. For example we can count the number flights in the data set using `r sourcecode2`. So that we print: ***There are `r length(unique(arctas_data$flight))` flights in the dataset***.

* Add a sentence to your markdown document that includes the number of flights in the data set.

---

## Appearance

The appearance of your **R Markdown** document can be formatted.

```
e.g. surround text with:
 * to make italic
 ** to make bold
 $ to add latex style equations
 [text](url) to add a web link to your markdown document
```

* add some bold and italic text to your markdown document
* add a link to the class github page

---

## Tables

There are multiple packages for adding tables to *R Markdown* documents for example *knitr*.

* add the code chunk to below to your document.

```{r}
  library(knitr)
  kable(arctas_data[1:5,1:4], caption = "Table caption")
```

---

# Data manipulation

---

## Packages

We'll use some of the **tidyverse** packages to manipulate our data:

You can include core **tidyverse** packages using:

```{r}
  library(tidyverse)
  library(magrittr)
```

Or you can include them explicitly:

```{r}
  library(dplyr)
  library(tidyr)
  library(purrr)
```

---

## Viewing data

* You can view data in **R Studio** using the `View()` function.

* You can use the `glimpse()` function to print a condensed summary of your data.

* Data of class `tbl` display in a convenient manner when the object is passed to the console

```
 * Use the function class(arctas_data) to dertermine the class of the arctas_data object

 * Use View(arctas_data) to display the data in R Studio

 * Use glimpse(arctas_data) to look at your data in the console
```

---

## Reshape data

Inevitably you will encounter data in inconvenient layouts. The `tidyr` package contains tools for reshaping your data.

* the `gather()` function converts data to a longer format

```
 * Add the code chunk below to your markdown
```

```{r}
  data_longer <- arctas_data %>%
                 select(flight, utc, no, noy, o3, no2_ncar, no2_ucb) %>%
                 gather("var", "val", 3:7)
```

```
 * What is the %>% doing? Trying typing ?"%>%" into the console.

 * Use the head() function to look at the first 5 lines of the data object
```

* the `spread()` function converts data to a wider format

```
 * Add the code chunk below to your markdown
```

```{r}
  data_wider <- data_longer %>%
                spread(var, val)
```

---

## Select data

Often you'll want to select a subset of your data. If your data is in a table you'll likely want to select by rows or columns. The `dplyr` package contains the tools you'll need to do this:

* the filter function allows you to select rows based on logical criteria for the values they contain

For example we can filter by column values:

```{r}
  data_filtered <- arctas_data %>%
                   filter(!is.na(no)) %>%
                   filter(flight == 4) %>%
                   filter(no > quantile(no, 0.05), no < quantile(no, 0.95))
```

```
  * what occurs at each filter step?
```

The `|` character is the logical `OR` operator e.g. `a == b | a == c`.

```
  Use the OR operator to select flight numbers 7, 11 and 21 from the dataset
```

* the select function allows you to select based on properties of the column names

For example:

```{r}
  data_select <- arctas_data %>%
                 select(flight, utc, no, noy, o3, no2_ncar, no2_ucb)
```

```
 * try selecting a different set of columns
```

---

## Combine data

Often data sets need to be combined. The `dplyr` package contains various functions function types for combining data.

* Perhaps the most simple case is adding rows or columns to a data set. You can use `bind_rows()` to add new rows and `bind_cols()` to add columns.

For example to combine data with the same column names:

```{r}
  flight_3  <- arctas_data %>%
               filter(flight == 3)

  flight_4 <-  arctas_data %>%
               filter(flight == 4)

  flight_3_4 <- bind_rows(flight_3, flight_4)
```

For example to combine data with the same number of rows:

```{r}
  data_n <- arctas_data %>%
            select(starts_with("n"))

  data_a <- arctas_data %>%
            select(starts_with("a"))
  
  data_n_a <- bind_cols(data_n, data_a)
```

* To join data by a common variable or variables use a `_join` function.

For example some notes pertaining to given flights

```{r}
  notes <- data_frame(flight = unique(arctas_data$flight)) %>%
           mutate(note = sample(letters, 22))

  data_with_note <- left_join(arctas_data, notes, by = "flight")
```

---

## New variables

The `mutate` functions in the `dplyr` can be used to create or modify columns in a data frame.

* The `mutate()` function can create a new variable like so:

* The `mutate_all()` function can be used to operate on all columns in a data frame.

* The `mutate_if()` function can be used to operate on columns that meet certain criteria.

---

## Grouping and summarizing data

The `group_by()` and `summarize` functions are useful for data reduction. Often you'll want to calculate statistics by groups within your data

* The `group_by()` function groups data by one or more variables. The data will look the same, but the results of subsequent operations on the data will be applied by groups.

* We can use `summarize()` to calculate statistics by group for a given column (variable)

* We can use `summarize_all()` to calculate the statistic for all the columns in a data frame.

* The data can be ungrouped using the `ungroup()` function.

```
 * group the data by flight and calculate the mean and standard deviation of the ozone mixing ratio
```

```{r}
  data_summary_o3 <- arctas_data %>%
                  group_by(flight) %>%
                  summarise(mean_o3 = mean(o3, na.rm = TRUE),
                            sd_o3 = sd(o3, na.rm = TRUE))
```

```
 * You can operate on all the columns
```
```{r}
  data_summary_all <- arctas_data %>%
                  group_by(flight) %>%
                  summarise_all(funs(mean_val = mean), na.rm = TRUE)
```

---

# Data visualization

We'll look at three related packages for visualizing data. The `ggplot2` package can be used to produce customized plots. The `ggmap` package extends `ggplot` for use with spatial data. The `plotly` package can be used to create interactive graphics.

* include the data visualization libraries in your markdown document:

```{r}
  library(ggplot2)
  library(ggmap)
  library(plotly)
  library(tidyverse)
  library(ggthemes)
```

---

## ggplot2

As a minimum, each `ggplot` graphic requires at least three elements:

* data
* aesthetic
* layer (geom)

Let's start with a time series plot.

```{r basics,}
  ggplot(data = filter(arctas_data,
                       flight == 10,
                       !is.na(o3)),
                mapping = aes(x = utc, y = o3)) +
  geom_point()
```

Let's plot multiple pollutants

* First convert the data to long format

```{r}
  data_plot <- arctas_data %>% 
               filter(flight == 10,
                      !is.na(o3)) %>%
               select(flight, utc, o3, no, no2_ncar, no3) %>%
               gather("var", "val", 3:6)
```

* Now we can plot the data

```{r,}
  ggplot(data_plot, aes(utc, val, color = var)) +
  geom_point() +
  scale_y_log10()
```

* All can control detailed aspects of you graphic's appearance using the `theme()` function. A quick way to change appearance is to use a theme function:

```
  * try applying the following themes to your plot:
```

```{r, eval = FALSE}
  + theme_economist()
  + theme_excel()     # as long as you promise never to use it again
  + theme_tufte()
```

 * It can be useful to separate into small-multiples (facets)
 
```{r, fig.width= 12, fig.height=20}
  data_plot <- arctas_data %>% 
               select(flight, utc, o3) %>%
               filter(!is.na(o3))

  ggplot(data_plot, aes(utc, o3)) +
  geom_point() +
  facet_wrap(~flight, ncol = 3)
```

---

## spatial data (ggmap)

The `ggmap` package builds on `ggplot` to map spatial data.

```{r}
  library(ggmap)
  library(ggplot2)
```

```
  * start by plotting a map of your favorite location:
```

```{r, eval = FALSE}
  qmap('favorite location', zoom = 13) 
```

```
  * now plot the arctas flight tracks on a world map
  * prepare the data
  * extract world map information
  * use ggplot to create a plot object
  * view the plot
```

* convert the longitude units:

```{r}
  data <- mutate(arctas_data,
                 longitude = ((longitude + 180) %% 360) - 180)
```

* a quick way to get some map data

```{r}
  map_world <- borders("world", colour="gray50", fill= "gray50")
```

* add flight track and plot

```{r,}
  map <- ggplot() +
         map_world +
         geom_point(aes(x = longitude, y = latitude, colour = o3),
                    size = 1,
                    data = data) +
         scale_colour_gradientn(colours = terrain.colors(10))

  map
```

* The `ggmaps` package contains tools for plotting spatial maps with greater control:

```
  * extract a single flight
```

```{r}
  data <- filter(arctas_data, flight == 8) %>%
          mutate(longitude = ((longitude + 180) %% 360) - 180)
```

```
  * calculate the required zoom
```

```{r}
  zoom <- calc_zoom(longitude, latitude, data = data) - 1
```

```
  * download map data
```

```{r}
  map <- get_map(location = c(lon = mean(data$longitude),
                               lat = mean(data$latitude)),
                 zoom = zoom,
                 maptype = "satellite",
                 source = "google")
```

```
  * combine the map with the flight track
```

```{r}
  map <- ggmap(map) +
         geom_path(aes(x = longitude, y = latitude, colour = o3),
                   size = 2,
                   data = data) +
          scale_colour_gradientn(colours = terrain.colors(10))
```

---

## Interactive maps

The [plotly](https://plot.ly/r/) package allows you to create interactive graphics.

```{r}
  library(plotly)
```

It works simply with `ggplot` objects:

* use the `ggplotly()` function to convert a `ggplot` object to an interactive `plotly` graphic.

```{r}
  ggplotly(map)
```
