---
title: "Week 2 - Dataframe manipulation"
author: "Nicholas Good (n.good@colostate.edu)"
output:
  rmarkdown::html_document:
    toc: true
    toc_float: true
    theme: yeti
---

```{r global_options, include=FALSE}
  knitr::opts_chunk$set(echo=TRUE, warning=FALSE, message=FALSE, results = 'hide')
```

---

# Getting started

First we'll need the `dplyr`, `readr` and `tibble` packages from `tidyverse`:

```{r}
 library(tidyverse)
```

Second we'll load in some data:

* What does the `[2]` suffix to the `files` object achieve?

```{r}
  files <- list.files("../data/w2_data", full.names = TRUE, pattern = "R14.ict$")

  f_1 <- read_csv(files[1], col_types = cols(), skip = 387)
  
  f_2 <- read_csv(files[2], col_types = cols(), skip = 387)
```

---

# Renaming

We'll look at two ways to rename columns.

## All the columns

You can use the `colnames()` function to return a character vector of column names:

```{r}
  names_f1 <- colnames(f_1)
```

We can then use the `names()` functions to modify the column names of an object:

```{r}
  f_lower <- f_1

  names(f_lower) <- tolower(colnames(f_lower))
```

## Individual columns

The `rename` function can be used to rename columns:

```{r}
  f_1_renamed <- rename(f_lower, time_utc = utc,
                                 julien_day = jday)
```

---

# New columns

The `mutate()` function creates a new column:

```{r}
  f_1_newcol <- mutate(f_lower, t_oc = temperature - 273.15)
```

---

# Filtering

The `filter()` function removes rows according to some condition:

```{r}
  f_1_filtered <- filter(f_lower, utc > 50000)
```

The `select()` function can be used to grab columns by name:

```{r}
  f_1_selected <- select(f_lower, utc, noy)
```

You can also the the `-` sign to drop columns:

```{r}
  f_1_selected <- select(f_lower, -utc, -noy)
```

---

# Aggregating

You can operate on groups of data by first grouping data using the `group_by()` function. Note: you can group `by` multiple variables.

```{r}
  f_1_flights <- group_by(f_lower, flight)
```

Once the data is grouped you can use a `window` function to performing a calculation and return the value for each group:

E.g. here we calculate the mean `fine_aerosol_sulfate` by `flight`.

```{r}
  f_1_mean_sulfate <- summarize(f_1_flights,
                                fine_aerosol_sulfate =
                                mean(fine_aerosol_sulfate, na.rm = TRUE))
```

---

# Merging

There are a number of `dplyr` functions for merging data. Two common tasks are combining rows of data with the same columns and adding columns to a data set.

* You can use the `bind_rows()` function to merge dataframes with identical column names:

```{r}
 f_data <- bind_rows(f_1, f_2)
```

* You can use the `_join()` functions to merge data sets by a common column or columns:

```{r}
  notes <- data_frame(FLIGHT = c(4, 5), note = c("note_1", "note2"))

  f_data_notes <- left_join(notes, f_data, by = "FLIGHT")
```

---

# Putting it all together

You can use the `%>%` operator from the `magrittr` package to make your code more readable:

```{r}
 f_all <- bind_rows(f_1, f_2) %>%
          mutate_all(funs(na_if(., -999999999))) %>%
          mutate_all(funs(na_if(., -888888888))) %>%
          set_names(tolower(names(.))) %>%
          group_by(flight) %>%
          summarise_all(mean, na.rm = TRUE)
```

There are also `%<>%`, `%$`, and `%T>%` operators in the `magrittr` package

---

# Exercise 2

1. Load the  `O3`, `NO`, `NO2`, and `NOY` data from the `flt20130603` data set into separate objects.

Hint: check the file types and starting rows.

2. Use the `left_join` function to combine the data into one `data_frame`.

Hint: which variable should you join by?

3. Convert missing values to `NA`.

Hint: use the `na_if()` function.

4. Rename the columns

---

