---
title: "Week 2 - Loading data"
author: "Nicholas Good (n.good@colostate.edu)"
output:
  rmarkdown::html_document:
    toc: true
    toc_float: true
    theme: yeti
---

```{r global_options, include=FALSE}
  knitr::opts_chunk$set(echo=TRUE, warning=FALSE, message=FALSE, results = 'hide')
```

---

# Set-up

* Let's organize our R session:

```{r}
 # 1. create a new R notebook for this exercise
 # 2. make sure you have this week's data folder downloaded from Google Drive
 # 3. check you working directory matches the project directory
```

---

# Text files

R includes numerous `read.` functions that can be used to load common files formats. The package `readr` also includes functions to load common files formats.

---

# Useful functions

Try running these functions, is there anything you need to change?

```{r}
  getwd()
  
  list.files(path = "../data/w2_data", full.names = TRUE)
  
  ?read.csv
```

---

## Loading a .csv

```{r}
 # store the file path in an object
  path <- "../data/w2_data/R for fire data analysis tutorials - Form Responses.csv"

 # use the read.csv function to load the file
 # what is each function argument doing?
  responses <- read.csv(file = path, header = TRUE, sep = ",")

 # explore the data you've just loaded using these functions
  head(responses)
  class(responses)
```

* Now try using the `readr` package

* Do you need to install the `readr` library? Try runnig `"readr" %in% rownames(installed.packages())` in your console.

* Do you need to update the `list.files` command?

* After loading the files, try typing the name of the file object into your console, what do you see?

* What class in the loaded object?

```{r}
  library(readr)

  path  <- "../data/w2_data/R for fire data analysis tutorials - Form Responses.csv"

  responses <- read_csv(file = path, col_names = TRUE)

  class(responses)
```

---

# Exercise 1

Loading an ARCTAS file

* Load the file `ARCTAS-mrg60-dc8_merge_20080409_R14.ict`.

* Open the file in your favorite text editor.

* What is the delimiter?

* Which load function should you use?

* Which line should you load first?

* Which libraries do you need?

---

# Messy files

You may come across files that are don't load nicely. For these you can use the `readr::read_lines` or the `base::readLines` function.

---

# Multiple files

Frequently you will want to load multiple files of the same format.

* We'll need to tell the function which files to load. You can use the `pattern =` argument to specify which files to include.

```{r}
  files <- list.files("../data/w2_data", full.names = TRUE, pattern = "R14.ict$")
```

* We can now use `lapply` to load each file nameds in the `files` object

* What class of object is created?

```{r}
  library(readr)

  multiple_files = lapply(files,
                   read_csv, skip = 387, col_types = cols())
```

* You can use the `dplyr` function `bind_rows` to create a single `data_frame`:

```{r}
  library(readr)
  library(dplyr)

  multiple_files = lapply(files,
                   read_csv, skip = 387, col_types = cols()) %>%
                   bind_rows()
```

---

# Other programs

If you would like to import data from other programming enviroments the are numerous R packages to facilitate this.

## Stats programs

The `haven` library has functions such as `read_sas()`, `read_sav()` and read `read_dta()` for SAS, SPSS, and STATA file types for example.

## Matlab

The `R.matlab` package has a `readMat()` function for loading .mat files.

---

# Saving data

## As R objects

You can R objects in a format R will recognizie or as text for example.

```{r}
  save(responses, file = "../data/responses.RData")

  saveRDS(responses, file = "../data/responses.RDS")
```

You can read R files like so:

```{r}
  load(file = "../data/responses.RData")

  responses_reloaded <- readRDS(responses, file = "../data/responses.RDS")
```

## As plain text

```{r}
  write.csv(responses, file = "../data/saved_responses_utils.csv")

  write_delim(responses, path = "../data/saved_responses_readr.csv", delim = "&")
```

---

